---
title: "Average Medicare charge for extracranial procedures"
author: "Brian High"
date: "5/16/2015"
output:
  html_document:
    keep_md: true
---

Introduction
------------

We are going to investigate the average Medicare charges for 
[extracranial procedures](https://www.cms.gov/icd10manual/fullcode_cms/P0054.html) 
for cities in Washington State, using 
[data](https://data.cms.gov/Medicare/Inpatient-Prospective-Payment-System-IPPS-Provider/97k6-zzx3) from the 
[U.S. Department of Health and Human Services, Centers for Medicare and Medicaid Services](http://www.cms.gov/).

Do the average Medicare covered charges for this procedure category vary from 
city to city in our state? 

This project serves as an example of using a single RMarkdown script to download 
data from a website, clean it up for analysis, and finally calculate 
some basic statistics and produce a plot. As the code and explanation are 
interwoven into the same executable document, this is an example of 
[literate programming](http://en.wikipedia.org/wiki/Literate_programming).

To reproduce this analysis, you can clone the 
[GitHub repository](https://github.com/brianhigh/medicare-ipps) for this
project into RStudio as a new project and execute the Rmd file with 
"Knit HTML" or Chunks -> Run All.

Prepare Environment
-------------------

Load the required packages.

```{r, echo=TRUE, message=FALSE}
# Load required packages. Install packages when necessary.
for (pkg in c("magrittr", "dplyr", "knitr", "ggplot2")) {
    if (! suppressWarnings(require(pkg, character.only=TRUE))) {
        install.packages(pkg, repos="http://cran.fhcrc.org", dependencies=TRUE)
        if (! suppressWarnings(require(pkg, character.only=TRUE)) ) {
            stop(paste0(c("Can't load package: ", pkg, "!"), collapse = ""))
        }
    }
}
```

Set up `knitr`.

```{r set_options, echo=TRUE, message=FALSE}
opts_chunk$set(tidy=FALSE, cache=FALSE)
```

Download the Data
-----------------

This script uses `download.file` to download the 
[data file](https://data.cms.gov/Medicare/Inpatient-Prospective-Payment-System-IPPS-Provider/97k6-zzx3) 
from [data.cms.gov](https://data.cms.gov).

```{r}
file <- "IPPS.csv"
url <- "http://data.cms.gov/api/views/97k6-zzx3/rows.csv?accessType=DOWNLOAD"
if (!file.exists(file)) { 
    download.file(url, file) 
}
```

Read and Subset the Data
------------------------

Read the CSV into a `data.frame` and `subset` by Washington Providers with a 
specific DRG Definition: '039 - EXTRACRANIAL PROCEDURES W/O CC/MCC'. Then 
`select` only the columns `Provider.City` and `Average.Covered.Charges`. Lastly, 
rename those columns to something a little shorter for convenience.

```{r}
ipps <- read.csv("IPPS.csv", header=T, stringsAsFactors=F)
ipps <- subset(ipps, Provider.State=='WA')
ipps <- subset(ipps, DRG.Definition=='039 - EXTRACRANIAL PROCEDURES W/O CC/MCC')
ipps <- select(ipps, Provider.City, Average.Covered.Charges)
colnames(ipps) <- c('City', 'Average.Charge')
```

```{r}
str(ipps)
```

Data Cleanup
------------

The `Average.Charge` has dollar signs and commas which we will remove using 
`sapply` and a custom function `clean_charge`.

```{r}
clean_charge <- function(charge) {
    as.character(charge) %>% gsub(pattern="[$,]", replacement="") %>% as.numeric
}

ipps$Average.Charge %<>% sapply(clean_charge)
```

Head and Tail
-------------

Now we'll take a look at the first and last lines of the data with `head` 
and `tail`.

```{r}
head(ipps)
tail(ipps)
```

Aggregate by City
-----------------

Since we have several cities with more than one Provider listed, group by 
`City` and calculate the mean for each group. Then sort by `Average.Charge` 
and apply this order to the `City` as factor levels.

```{r}
ipps <- aggregate(Average.Charge~City, ipps, mean)
ipps <- arrange(ipps, Average.Charge, City)
ipps$City <- factor(ipps$City, as.character(ipps$City))
```

Nicer Table
-----------

Format the table a little more nicely for our report using `kable`.

```{r}
kable(ipps, "html")
```

Bar Plot
--------

Finally, create a bar plot with `ggplot`.

```{r}
ggplot(ipps, aes(x=City, y=Average.Charge)) +
    geom_bar(stat="identity", colour="darkblue", fill="darkblue") +  
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ggtitle("Average covered Medicare charges for \nextracranial procedures in WA") +
    labs(x="City", y="Average Covered Charges (US$)")
```