---
title: "Average Medicare charge for extracranial procedures"
author: "Brian High"
date: "5/4/2015"
output:
  html_document:
    keep_md: true
---

Prepare Environment
--------------------

Prepare the environment by loading the required packages ...

```{r}
# Load packages
pkgs <- c("magrittr", "dplyr", "knitr", "ggplot2")
# install.packages(pkgs)       # Uncomment to install the packages if you need to.
invisible(suppressMessages(suppressWarnings(lapply(pkgs, require, character.only=T))))
```

... and setting up `knitr`.

```{r set_options, echo=TRUE, message=FALSE}
opts_chunk$set(tidy=FALSE, cache=TRUE)
```

Download the Data
-----------------

This script uses `download.file` to download the 
[data file](https://data.cms.gov/Medicare/Inpatient-Prospective-Payment-System-IPPS-Provider/97k6-zzx3) 
from [data.cms.gov](https://data.cms.gov).

```{r}
file <- "IPPS.csv"
url <- "http://data.cms.gov/api/views/97k6-zzx3/rows.csv?accessType=DOWNLOAD"
if (!file.exists(file)) { 
    download.file(url, file) 
}
```

Read and Subset the Data
------------------------

Read the CSV into a `data.frame` and `subset` by Washington Providers  
with a specific DRG Definition: '039 - EXTRACRANIAL PROCEDURES W/O CC/MCC'. Then
`select` only the columns `Provider.City` and `Average.Covered.Charges`. Lastly, 
rename those columns to something a little shorter for convenience.

```{r}
ipps <- read.csv("IPPS.csv", header=T, stringsAsFactors=F)
ipps <- subset(ipps, Provider.State=='WA')
ipps <- subset(ipps, DRG.Definition=='039 - EXTRACRANIAL PROCEDURES W/O CC/MCC')
ipps <- select(ipps, Provider.City, Average.Covered.Charges)
colnames(ipps) <- c('City', 'Average.Charge')
```

```{r}
str(ipps)
```

Data Cleanup
------------

The `Average.Charge` has dollar signs and commas which we will remove using 
`sapply` and a custom function `clean_charge`.

```{r}
clean_charge <- function(charge) {
    as.character(charge) %>% gsub(pattern="[$,]", replacement="") %>% as.numeric
}

ipps$Average.Charge %<>% sapply(clean_charge)
```

Head and Tail
-------------

Now we'll take a look at the first and last lines of the data with `head` 
and `tail`.

```{r}
head(ipps)
tail(ipps)
```

Aggregate by City
-----------------

Since we have several cities with more than one Provider listed, group by 
`City` and calculate the mean for each group. Then sort by `Average.Charge` 
and apply this order to the `City` as factor levels.

```{r}
ipps <- aggregate(Average.Charge~City, ipps, mean)
ipps <- arrange(ipps, Average.Charge, City)
ipps$City <- factor(ipps$City, as.character(ipps$City))
```

Nicer Table
-----------

Format the table a little more nicely for our report using `kable`.

```{r}
kable(ipps, "html")
```

Bar Plot
--------

Finally, create a bar plot with `ggplot`.

```{r}
ggplot(ipps, aes(x=City, y=Average.Charge)) +
    geom_bar(stat="identity", colour="black", fill="lightblue") +  
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    ggtitle("Average covered Medicare charges for \nextracranial procedures in WA") +
    labs(x="City", y="Average Covered Charges (US$)")
```