---
title: "Average Medicare charge for extracranial procedures"
author: "Brian High"
date: "5/5/2015"
output:
  html_document:
    keep_md: true
---

Introduction
------------

We are going to investigate the average Medicare charges for 
[extracranial procedures](https://www.cms.gov/icd10manual/fullcode_cms/P0054.html) 
for cities in Washington State, using 
[data](https://data.cms.gov/Medicare/Inpatient-Prospective-Payment-System-IPPS-Provider/97k6-zzx3) from the 
[U.S. Department of Health and Human Services, Centers for Medicare and Medicaid Services](http://www.cms.gov/).

Do the average Medicare covered charges for this procedure category vary from 
city to city in our state? 

To find out, we will use a dataset hosted and publically shared on the 
UW [SQLShare service using R](http://escience.washington.edu/get-help-now/accessing-sqlshare-r).

This project serves as an example of using a single RMarkdown script to access
data from a SQL database, clean it up for analysis, and finally calculate 
some basic statistics and produce a plot. As the code and explanation are 
interwoven into the same executable document, this is an example of 
[literate programming](http://en.wikipedia.org/wiki/Literate_programming).

To reproduce this analysis, you can clone the 
[GitHub repository](https://github.com/brianhigh/medicare-ipps) for this
project into RStudio as a new project and execute the Rmd file with 
"Knit HTML" or Chunks -> Run All.

SQLShare Config
---------------

As this script requires a SQLShare account and API key, check for the 
`config` file. If you don't have one, check the 
[SQLShare website](http://escience.washington.edu/get-help-now/accessing-sqlshare-r) 
for more information on how to create this file.

```{r}
# Check for sqlshare config file and abort if not found
stopifnot(file.exists("~/.sqlshare/config"))
```

Prepare Environment
-------------------

Prepare the environment by loading the required packages...

```{r}
# tryinstall() function
#     Conditionally install using install.packages()
#     Usage: tryinstall(c("package1", "package2", ...))
tryinstall <- function(p) {
    n <- p[!(p %in% installed.packages()[,"Package"])]
    if(length(n)) {
        install.packages(n, repos="http://cran.fhcrc.org", dependencies=TRUE)
    }
}

# You will need to load these packages
pkgs <- c("sqlshare", "magrittr", "dplyr", "knitr", "ggplot2")

# Attempt to load or install the packages using the tryinstall function
tryinstall(pkgs)

# ... or uncomment these next three lines to manually install packages.
# install.packages(
#     c("sqlshare", "magrittr", "dplyr", "knitr", "ggplot2"), 
#     repos="http://cran.fhcrc.org", dependencies=TRUE)

# Check that all packages will load
for (pkg in pkgs) suppressPackageStartupMessages(suppressWarnings(
  require(pkg, character.only=TRUE, quietly=TRUE)))
```

... and setting up `knitr`.

```{r set_options, echo=TRUE, message=FALSE}
opts_chunk$set(tidy=FALSE, cache=FALSE)
```

Query Database
--------------

Define our query and submit it to the SQLShare service.

```{r}
# Data source: 
# https://data.cms.gov/Medicare/Inpatient-Prospective-Payment-System-IPPS-Provider/97k6-zzx3
# https://sqlshare.escience.washington.edu/sqlshare/#s=query/high%40washington.edu/IPPS.csv
sql <- "SELECT [Provider City],[Average Covered Charges] 
        FROM [high@washington.edu].[table_IPPS.csv] 
        WHERE [Provider State] = 'WA' 
        AND [DRG Definition] = '039 - EXTRACRANIAL PROCEDURES W/O CC/MCC'"
ipps <- fetch.data.frame(sql)
```

The SQLShare package has returned our data as a `data.frame`. 

```{r}
str(ipps)
```

Data Cleanup
------------

The `Average.Covered.Charge` amounts are stored as a factor, not as a number. We 
should fix that, after removing the dollar signs and commas. First, let's rename 
the columns, then use `sapply` and a custom function `clean_charge` to do the 
rest.

```{r}
colnames(ipps) <- c('City', 'Average.Charge')

clean_charge <- function(charge) {
    as.character(charge) %>% gsub(pattern="[$,]", replacement="") %>% as.numeric
}

ipps$Average.Charge %<>% sapply(clean_charge)
```

Head and Tail
-------------

Now we'll take a look at the first and last lines of the data with `head` 
and `tail`.

```{r}
head(ipps)
tail(ipps)
```

Aggregate by City
-----------------

Since we have several cities with more than one Provider listed, group by 
`City` and calculate the mean for each group. Then sort by `Average.Charge`, 
making sure to also apply this order to the `City` factor levels.

```{r, }
ipps <- aggregate(Average.Charge~City, ipps, mean)
ipps <- arrange(ipps, Average.Charge, City)
ipps$City <- factor(ipps$City, as.character(ipps$City))
```

Nicer Table
-----------

Format the table a little more nicely for our report using `kable`.

```{r}
kable(ipps, "html")
```

Bar Plot
--------

Finally, create a bar plot with `ggplot`.

```{r}
ggplot(ipps, aes(x=City, y=Average.Charge)) +
    geom_bar(stat="identity", colour="darkblue", fill="darkblue") +  
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    ggtitle("Average covered Medicare charges for \nextracranial procedures in WA") +
    labs(x="City", y="Average Covered Charges (US$)")
```